---
title: "Why zztab2fig? A Practical Comparison with pander"
author: "Ronald G. Thomas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why zztab2fig? A Practical Comparison with pander}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)

has_latex <- nzchar(Sys.which("pdflatex")) && nzchar(Sys.which("pdfcrop"))

output_dir <- tempdir()

include_table <- function(filename) {
  pdf_path <- file.path(output_dir, paste0(filename, "_cropped.pdf"))
  if (!file.exists(pdf_path)) return(invisible(NULL))
  png_path <- sub("\\.pdf$", ".png", pdf_path)
  convert_pdf_to_png(pdf_path, png_path, dpi = 150)
  knitr::include_graphics(png_path)
}
```

## Introduction

R provides multiple packages for creating tables from data frames, each with
distinct design philosophies. This vignette compares **pander** and
**zztab2fig** through practical examples to help you choose the right tool.

**pander** is designed for quick, inline table generation in R Markdown
documents. It converts R objects to Markdown format, which Pandoc then renders
to HTML, PDF, or Word. This approach is fast and integrates seamlessly with
the R Markdown ecosystem.
**zztab2fig** takes a different approach: it generates standalone PDF tables
via LaTeX compilation, then crops them for inclusion in documents. This
produces publication-quality tables with features that Markdown cannot
express, such as footnotes, spanning headers, and decimal alignment.

The choice depends on your needs:

- **Quick reports and HTML output**: pander is simpler and faster
- **Journal submissions and formal publications**: zztab2fig provides the
  control and formatting features required

```{r libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(zztab2fig))
```


# Part 1: Basic Tabling

These examples show that both packages handle simple tables well. The
differences emerge in formatting control and output quality.

## Challenge 1: Simple Data Frame

**Task**: Display a basic data frame.

### pander Approach

```r
df <- mtcars[1:5, 1:4]
pander::pander(df)
```

**pander output** (Markdown, rendered by Pandoc):

```
-------------------------------------------
       &nbsp;         mpg   cyl   disp   hp
-------------------- ----- ----- ------ ----
   **Mazda RX4**      21     6    160   110

 **Mazda RX4 Wag**    21     6    160   110

  **Datsun 710**     22.8    4    108    93
-------------------------------------------
```

### zztab2fig Approach

```{r challenge1-zztab2fig, eval=has_latex}
df <- mtcars[1:5, 1:4]
t2f(df, filename = "basic_table", sub_dir = output_dir)
```

```{r challenge1-display, echo=FALSE, eval=has_latex, out.width="70%"}
include_table("basic_table")
```

**Key difference**: zztab2fig produces a PDF with professional typesetting
(booktabs rules, proper spacing). pander produces Markdown for inline display.


## Challenge 2: Column Alignment

**Task**: Left-align text columns, right-align numeric columns.

### pander

```r
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Score = c(95.5, 87.2, 91.8),
  Rank = c(1, 3, 2)
)
pander::pander(df, justify = c("left", "right", "right"))
```

### zztab2fig

```{r challenge2-zztab2fig, eval=has_latex}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Score = c(95.5, 87.2, 91.8),
  Rank = c(1, 3, 2)
)
t2f(df, filename = "aligned_table", sub_dir = output_dir,
    align = c("l", "r", "r"))
```

```{r challenge2-display, echo=FALSE, eval=has_latex, out.width="50%"}
include_table("aligned_table")
```

**Key difference**: Both handle L/C/R alignment. zztab2fig also supports
decimal alignment via siunitx (shown later).


## Challenge 3: Table with Caption

**Task**: Add a descriptive caption.

### pander

```r
df <- iris[1:5, ]
pander::pander(df, caption = "First Five Rows of Iris Dataset")
```

Produces caption below the table with "Table:" prefix.

### zztab2fig

```{r challenge3-zztab2fig, eval=has_latex}
df <- iris[1:5, ]
t2f(df,
    filename = "iris_sample",
    sub_dir = output_dir,
    caption = "First Five Rows of Iris Dataset",
    label = "tab:iris")
```

```{r challenge3-display, echo=FALSE, eval=has_latex, out.width="70%"}
include_table("iris_sample")
```

**Key difference**: zztab2fig adds LaTeX cross-reference labels for use with
`\ref{tab:iris}` in documents. pander has no equivalent.


## Challenge 4: Regression Model Output

**Task**: Display coefficients from a linear model.

### pander

```r
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)
pander::pander(model)
```

Produces a markdown table of coefficients.

### zztab2fig

```{r challenge4-zztab2fig, eval=has_latex}
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)
t2f(model,
    filename = "regression",
    sub_dir = output_dir,
    include = c("estimate", "std.error", "p.value"),
    caption = "Linear Regression Results")
```

```{r challenge4-display, echo=FALSE, eval=has_latex, out.width="60%"}
include_table("regression")
```

**Key difference**: Both handle `lm` objects. zztab2fig provides S3 methods
for selecting which statistics to display (`include` parameter) and supports
additional object types (glm, anova, htest).


# Part 2: Advanced Features

These challenges demonstrate capabilities unique to zztab2fig. pander cannot
replicate these without manual LaTeX coding.


## Challenge 5: Table Footnotes

**Task**: Add footnotes explaining abbreviations and significance levels.

### pander

```r
df <- data.frame(
  Variable = c("BMI*", "SBP", "DBP"),
  Mean = c(27.5, 142.3, 88.2),
  SD = c(4.2, 18.5, 11.3)
)
pander::pander(df, caption = "Clinical Measurements")
# Footnotes must be added as separate text below the table
```

**Limitation**: pander has no table footnote support. Notes must be added
manually as prose.

### zztab2fig

```{r challenge5-zztab2fig, eval=has_latex}
df <- data.frame(
  Variable = c("BMI", "SBP", "DBP"),
  Mean = c(27.5, 142.3, 88.2),
  SD = c(4.2, 18.5, 11.3)
)

df$Variable[1] <- t2f_mark("BMI", 1, "symbol")

fn <- t2f_footnote(
  general = "SBP = Systolic Blood Pressure; DBP = Diastolic Blood Pressure.",
  symbol = "Body Mass Index (kg/m2).",
  threeparttable = TRUE
)

t2f(df,
    filename = "clinical_footnotes",
    sub_dir = output_dir,
    caption = "Clinical Measurements",
    footnote = fn)
```

```{r challenge5-display, echo=FALSE, eval=has_latex, out.width="50%"}
include_table("clinical_footnotes")
```

**Advantage**: zztab2fig produces proper LaTeX footnotes using threeparttable,
with support for general notes, numbered notes, and symbol notes.


## Challenge 6: Spanning Column Headers

**Task**: Group related columns under a common header (e.g., "Treatment" and
"Control" groups).

### pander

```r
# pander cannot create spanning headers
# Best approximation: rename columns to indicate grouping
names(df) <- c("Outcome", "Treatment Mean", "Treatment SD",
               "Control Mean", "Control SD")
pander::pander(df)
```

**Limitation**: pander cannot create multi-column headers. Column names must
encode the grouping.

### zztab2fig

```{r challenge6-zztab2fig, eval=has_latex}
df <- data.frame(
  Outcome = c("Score A", "Score B"),
  T_Mean = c(45.2, 38.7),
  T_SD = c(8.3, 6.2),
  C_Mean = c(42.1, 37.9),
  C_SD = c(7.9, 5.8)
)
names(df) <- c("Outcome", "Mean", "SD", "Mean", "SD")

hdr <- t2f_header_above(
  " " = 1,
  "Treatment" = 2,
  "Control" = 2
)

t2f(df,
    filename = "spanning_header",
    sub_dir = output_dir,
    caption = "Outcomes by Group",
    header_above = hdr)
```

```{r challenge6-display, echo=FALSE, eval=has_latex, out.width="60%"}
include_table("spanning_header")
```

**Advantage**: zztab2fig creates proper LaTeX multicolumn headers with
cmidrule separators, essential for clinical trial and research tables.


## Challenge 7: Multi-Row Cells (Hierarchical Data)

**Task**: Merge repeated values vertically to show data hierarchy.

### pander

```r
df <- data.frame(
  Category = c("Treatment", "Treatment", "Control", "Control"),
  Subgroup = c("Male", "Female", "Male", "Female"),
  N = c(45, 52, 48, 49),
  Response = c("72%", "68%", "45%", "42%")
)
# pander displays repeated values without merging
pander::pander(df)
```

**Limitation**: pander cannot merge cells vertically.

### zztab2fig

```{r challenge7-zztab2fig, eval=has_latex}
df <- data.frame(
  Category = c("Treatment", "Treatment", "Control", "Control"),
  Subgroup = c("Male", "Female", "Male", "Female"),
  N = c(45, 52, 48, 49),
  Response = c("72%", "68%", "45%", "42%")
)

t2f(df,
    filename = "multirow",
    sub_dir = output_dir,
    caption = "Response by Group and Sex",
    collapse_rows = t2f_collapse_rows(
      columns = 1,
      valign = "middle",
      latex_hline = "major"
    ))
```

```{r challenge7-display, echo=FALSE, eval=has_latex, out.width="55%"}
include_table("multirow")
```

**Advantage**: zztab2fig automatically collapses repeated values using LaTeX
multirow, with configurable vertical alignment and horizontal rules.


## Challenge 8: Decimal Point Alignment

**Task**: Align numbers on the decimal point for easy visual comparison.

### pander

```r
df <- data.frame(
  Item = c("A", "B", "C"),
  Value1 = c(1.5, 123.45, 12.345),
  Value2 = c(0.001, 10.1, 1000.01)
)
# pander uses right-alignment (no decimal alignment)
pander::pander(df, justify = c("left", "right", "right"))
```

**Limitation**: pander only supports left/center/right alignment, not decimal
alignment.

### zztab2fig

```{r challenge8-zztab2fig, eval=has_latex}
df <- data.frame(
  Item = c("A", "B", "C"),
  Value1 = c(1.5, 123.45, 12.345),
  Value2 = c(0.001, 10.1, 1000.01)
)

t2f(df,
    filename = "decimal_aligned",
    sub_dir = output_dir,
    align = list(
      "l",
      t2f_siunitx(table_format = "3.3"),
      t2f_siunitx(table_format = "4.2")
    ))
```

```{r challenge8-display, echo=FALSE, eval=has_latex, out.width="45%"}
include_table("decimal_aligned")
```

**Advantage**: zztab2fig uses siunitx for true decimal alignment, making
numeric comparisons significantly easier to read.


## Challenge 9: Journal-Specific Themes

**Task**: Format a table for NEJM (New England Journal of Medicine)
submission.

### pander

```r
# pander has no journal-specific themes
# Manual formatting required for each style guide
pander::panderOptions("table.style", "rmarkdown")
pander::pander(df, caption = "Baseline Characteristics")
```

**Limitation**: pander provides no built-in journal themes.

### zztab2fig

```{r challenge9-zztab2fig, eval=has_latex}
df <- data.frame(
  Characteristic = c("Age, years", "Male sex, n (%)", "BMI, kg/m2"),
  Treatment = c("65.2 (8.4)", "142 (58%)", "27.3 (4.1)"),
  Placebo = c("64.8 (8.1)", "138 (56%)", "27.1 (3.9)")
)

t2f(df,
    filename = "nejm_table",
    sub_dir = output_dir,
    caption = "Baseline Characteristics",
    theme = "nejm")
```

```{r challenge9-display, echo=FALSE, eval=has_latex, out.width="55%"}
include_table("nejm_table")
```

**Advantage**: zztab2fig includes built-in themes for NEJM, APA, Nature, and
a minimal style. Each theme sets appropriate fonts, spacing, and rules.


## Challenge 10: Model Comparison Table

**Task**: Display multiple regression models side-by-side with significance
stars.

### pander

```r
m1 <- lm(mpg ~ cyl, data = mtcars)
m2 <- lm(mpg ~ cyl + hp, data = mtcars)
m3 <- lm(mpg ~ cyl + hp + wt, data = mtcars)

# pander displays one model at a time
pander::pander(m1)
pander::pander(m2)
pander::pander(m3)
# Side-by-side comparison requires manual table construction
```

**Limitation**: pander cannot create comparative model tables automatically.

### zztab2fig

```{r challenge10-zztab2fig, eval=has_latex}
m1 <- lm(mpg ~ cyl, data = mtcars)
m2 <- lm(mpg ~ cyl + hp, data = mtcars)
m3 <- lm(mpg ~ cyl + hp + wt, data = mtcars)

t2f_regression(
  Model1 = m1,
  Model2 = m2,
  Model3 = m3,
  stars = TRUE,
  filename = "model_comparison",
  sub_dir = output_dir
)
```

```{r challenge10-display, echo=FALSE, eval=has_latex, out.width="70%"}
include_table("model_comparison")
```

**Advantage**: `t2f_regression()` automatically aligns terms across models,
adds significance stars, and includes model statistics (N, R-squared).


## Challenge 11: Complex Combined Features

**Task**: Create a publication-quality table with footnotes, spanning headers,
multi-row cells, and NEJM styling.

### pander

```r
# This combination cannot be achieved in pander
# Requires manual LaTeX coding
```

### zztab2fig

```{r challenge11-zztab2fig, eval=has_latex}
df <- data.frame(
  Endpoint = c("Primary", "Primary", "Secondary", "Secondary"),
  Timepoint = c("Week 26", "Week 52", "Week 26", "Week 52"),
  N = c(245, 232, 245, 232),
  Difference = c(-0.42, -0.58, -0.28, -0.35),
  P = c("0.008", "0.002", "0.045", "0.018")
)

df$Difference <- sapply(seq_len(nrow(df)), function(i) {
  p <- as.numeric(df$P[i])
  if (p < 0.01) t2f_mark(as.character(df$Difference[i]), 2, "symbol")
  else if (p < 0.05) t2f_mark(as.character(df$Difference[i]), 1, "symbol")
  else as.character(df$Difference[i])
})

hdr <- t2f_header_above(" " = 2, "Results" = 3)

fn <- t2f_footnote(
  general = "Negative values favor treatment.",
  symbol = c("p < 0.05", "p < 0.01")
)

t2f(df,
    filename = "complex_table",
    sub_dir = output_dir,
    caption = "Efficacy Results by Endpoint and Timepoint",
    caption_short = "Efficacy Results",
    header_above = hdr,
    collapse_rows = t2f_collapse_rows(1, valign = "top"),
    footnote = fn,
    theme = "nejm")
```

```{r challenge11-display, echo=FALSE, eval=has_latex, out.width="65%"}
include_table("complex_table")
```


# Summary

## Feature Comparison

| Feature | pander | zztab2fig |
|:--------|:------:|:---------:|
| Basic data frames | Yes | Yes |
| Column alignment (L/C/R) | Yes | Yes |
| Decimal alignment | No | Yes (siunitx) |
| Captions | Yes | Yes + short captions |
| Cross-reference labels | No | Yes |
| Table footnotes | No | Yes (4 notation types) |
| Spanning headers | No | Yes |
| Multi-row cells | No | Yes |
| Multi-page tables | No | Yes (longtable) |
| Journal themes | No | Yes (NEJM, APA, Nature) |
| Model comparison | No | Yes |
| Statistical object S3 methods | Partial | Yes (lm, glm, anova, htest) |
| Output formats | Markdown | PDF, PNG, SVG, TEX |

## When to Use Each Package

### Choose pander when:

- Creating R Markdown documents for HTML or Word output
- Need quick, inline display of many R object types
- Document will be processed through Pandoc
- Simple tables with basic formatting are sufficient
- Rapid iteration is more important than typographic quality

### Choose zztab2fig when:

- Creating tables for journal submission
- Need footnotes, spanning headers, or multi-row cells
- Working with LaTeX documents directly
- Require decimal alignment for numeric comparison
- Want consistent journal-specific styling (NEJM, APA, Nature)
- Need cropped PDF tables for inclusion in external documents
- Building model comparison tables

## v0.2.0 Features Used in This Vignette

This vignette demonstrates several features introduced in zztab2fig v0.2.0:

- **S3 methods**: `t2f.lm()` for linear models
- **Theme system**: Built-in NEJM theme (`theme = "nejm"`)
- **t2f_regression()**: Side-by-side model comparison
- **t2f_footnote()**: Structured footnotes with multiple notation types
- **t2f_header_above()**: Spanning column headers
- **t2f_collapse_rows()**: Multi-row cell merging
- **t2f_siunitx()**: Decimal alignment
- **t2f_mark()**: Footnote markers in cells
- **caption_short**: Short captions for List of Tables

For complete documentation of these features, see the "Advanced Features" and
"Object Types and Themes" vignettes.
