---
title: "Tabling Challenges: pander vs zztab2fig"
author: "Ronald G. Thomas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tabling Challenges: pander vs zztab2fig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
This vignette presents a series of tabling challenges of varying difficulty,
showing solutions in both pander and zztab2fig. Each challenge demonstrates
the relative strengths and approaches of each package.

## Setup

```{r libraries}
library(pander)
library(zztab2fig)
```

---

# Standard Difficulty Challenges

## Challenge 1: Basic Data Frame

**Task**: Display a simple data frame with default formatting.

### pander Solution

```{r challenge1-pander}
df <- mtcars[1:5, 1:4]

pander(df)
```

**Output** (Markdown):
```
-------------------------------------------
       &nbsp;         mpg   cyl   disp   hp
-------------------- ----- ----- ------ ----
   **Mazda RX4**      21     6    160   110

 **Mazda RX4 Wag**    21     6    160   110

  **Datsun 710**     22.8    4    108    93

 **Hornet 4 Drive**  21.4    6    258   110

**Hornet Sportabout** 18.7   8    360   175
-------------------------------------------
```

### zztab2fig Solution

```{r challenge1-zztab2fig}
df <- mtcars[1:5, 1:4]

t2f(df, filename = "basic_table")
```

**Output**: Cropped PDF with LaTeX booktabs formatting.

**Comparison**: pander outputs markdown for inline display; zztab2fig produces
publication-ready PDF files.

---

## Challenge 2: Custom Column Alignment

**Task**: Create a table with left-aligned text and right-aligned numbers.

### pander Solution

```{r challenge2-pander}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Score = c(95.5, 87.2, 91.8),
  Rank = c(1, 3, 2)
)

pander(df, justify = c("left", "right", "right"))
```

### zztab2fig Solution

```{r challenge2-zztab2fig}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Score = c(95.5, 87.2, 91.8),
  Rank = c(1, 3, 2)
)

# Auto-detection (default)
t2f(df, filename = "aligned_table")

# Or explicit
t2f(df, filename = "aligned_table", align = c("l", "r", "r"))
```

**Comparison**: Both handle basic alignment. zztab2fig auto-detects based on
column type.

---

## Challenge 3: Table with Caption

**Task**: Add a caption to the table.

### pander Solution

```{r challenge3-pander}
df <- iris[1:5, ]

pander(df, caption = "First Five Rows of Iris Dataset")
```

**Output**: Caption appears below table with "Table:" prefix.

### zztab2fig Solution

```{r challenge3-zztab2fig}
df <- iris[1:5, ]

t2f(df,
    filename = "iris_sample",
    caption = "First Five Rows of Iris Dataset",
    label = "tab:iris")
```

**Output**: LaTeX `\caption{}` with optional `\label{}` for cross-referencing.

**Comparison**: zztab2fig adds cross-reference labels; pander does not.

---

## Challenge 4: Regression Model Output

**Task**: Display linear regression coefficients.

### pander Solution

```{r challenge4-pander}
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)

pander(model)
```

**Output**:
```
---------------------------------------------------------------
     &nbsp;        Estimate   Std. Error   t value   Pr(>|t|)
----------------- ---------- ------------ --------- ----------
 **(Intercept)**    38.75       1.787       21.69    3.043e-18

     **cyl**       -0.9416      0.5509     -1.709     0.0985

      **hp**       -0.01804    0.01188     -1.519     0.1400

      **wt**        -3.167      0.7406     -4.276    0.0002073
---------------------------------------------------------------
```

### zztab2fig Solution

```{r challenge4-zztab2fig}
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)

t2f(model,
    filename = "regression",
    include = c("estimate", "std.error", "p.value"),
    caption = "Linear Regression Results")
```

**Output**: Formatted coefficient table with selected statistics.

**Comparison**: Both handle lm objects natively. zztab2fig allows selecting
which statistics to include.

---

## Challenge 5: Highlighting Cells

**Task**: Emphasize specific cells in a table.

### pander Solution

```{r challenge5-pander}
df <- mtcars[1:5, 1:4]

# Emphasize row 2, column 1
pander(df, emphasize.strong.cells = which(df == df[2,1], arr.ind = TRUE))

# Or emphasize entire rows/columns
pander(df, emphasize.strong.rows = 2, emphasize.italics.cols = 3)
```

### zztab2fig Solution

```{r challenge5-zztab2fig}
df <- mtcars[1:5, 1:4]

# Bold a column
t2f(df,
    filename = "emphasized",
    formatting = list(t2f_bold_col(1)))

# Conditional highlighting
fmt <- t2f_highlight(
  condition = function(x) as.numeric(x) > 20,
  background = "yellow!20"
)
```

**Comparison**: pander uses row/column indices; zztab2fig supports conditional
formatting with LaTeX colors.

---

# Advanced Challenges

## Challenge 6: Table Footnotes

**Task**: Add footnotes explaining abbreviations and significance levels.

### pander Solution

```{r challenge6-pander}
df <- data.frame(
  Variable = c("BMI*", "SBP", "DBP"),
  Mean = c(27.5, 142.3, 88.2),
  SD = c(4.2, 18.5, 11.3)
)

pander(df, caption = "Clinical Measurements")
# Footnote must be added as separate text outside the table
cat("\n\n*BMI = Body Mass Index; SBP = Systolic Blood Pressure;
DBP = Diastolic Blood Pressure\n")
```

**Limitation**: pander has no native table footnote support. Footnotes must be
added manually as separate text.

### zztab2fig Solution

```{r challenge6-zztab2fig}
df <- data.frame(
  Variable = c("BMI", "SBP", "DBP"),
  Mean = c(27.5, 142.3, 88.2),
  SD = c(4.2, 18.5, 11.3)
)

# Add footnote marker to cell
df$Variable[1] <- t2f_mark("BMI", 1, "symbol")

fn <- t2f_footnote(
  general = "SBP = Systolic Blood Pressure; DBP = Diastolic Blood Pressure.",
  symbol = "Body Mass Index (kg/m²).",
  threeparttable = TRUE
)

t2f(df,
    filename = "clinical_footnotes",
    caption = "Clinical Measurements",
    footnote = fn)
```

**Advantage**: zztab2fig produces proper LaTeX footnotes within the table
environment using threeparttable.

---

## Challenge 7: Spanning Column Headers

**Task**: Create grouped column headers (e.g., "Treatment" spanning two
columns).

### pander Solution

```{r challenge7-pander}
df <- data.frame(
  Outcome = c("Score A", "Score B"),
  T_Mean = c(45.2, 38.7),
  T_SD = c(8.3, 6.2),
  C_Mean = c(42.1, 37.9),
  C_SD = c(7.9, 5.8)
)

# pander cannot create spanning headers
# Best approximation: rename columns
names(df) <- c("Outcome", "Treatment Mean", "Treatment SD",
               "Control Mean", "Control SD")
pander(df)
```

**Limitation**: pander does not support spanning (multi-column) headers.

### zztab2fig Solution

```{r challenge7-zztab2fig}
df <- data.frame(
  Outcome = c("Score A", "Score B"),
  T_Mean = c(45.2, 38.7),
  T_SD = c(8.3, 6.2),
  C_Mean = c(42.1, 37.9),
  C_SD = c(7.9, 5.8)
)
names(df) <- c("Outcome", "Mean", "SD", "Mean", "SD")

hdr <- t2f_header_above(
  " " = 1,
  "Treatment" = 2,
  "Control" = 2
)

t2f(df,
    filename = "spanning_header",
    caption = "Outcomes by Group",
    header_above = hdr)
```

**Advantage**: zztab2fig creates proper LaTeX multicolumn headers with cmidrule
separators.

---

## Challenge 8: Multi-Row Cells (Row Spanning)

**Task**: Merge cells vertically for hierarchical data.

### pander Solution

```{r challenge8-pander}
df <- data.frame(
  Category = c("Treatment", "Treatment", "Control", "Control"),
  Subgroup = c("Male", "Female", "Male", "Female"),
  N = c(45, 52, 48, 49),
  Response = c("72%", "68%", "45%", "42%")
)

# pander cannot merge rows - display as-is with repeated values
pander(df)
```

**Limitation**: pander does not support multi-row (vertically merged) cells.

### zztab2fig Solution

```{r challenge8-zztab2fig}
df <- data.frame(
  Category = c("Treatment", "Treatment", "Control", "Control"),
  Subgroup = c("Male", "Female", "Male", "Female"),
  N = c(45, 52, 48, 49),
  Response = c("72%", "68%", "45%", "42%")
)

t2f(df,
    filename = "multirow",
    caption = "Response by Group and Sex",
    collapse_rows = t2f_collapse_rows(
      columns = 1,
      valign = "middle",
      latex_hline = "major"
    ))
```

**Advantage**: zztab2fig automatically merges repeated values using LaTeX
multirow.

---

## Challenge 9: Decimal Point Alignment

**Task**: Align numbers on the decimal point for easy comparison.

### pander Solution

```{r challenge9-pander}
df <- data.frame(
  Item = c("A", "B", "C"),
  Value1 = c(1.5, 123.45, 12.345),
  Value2 = c(0.001, 10.1, 1000.01)
)

# pander uses right-alignment for numbers
# No decimal alignment available
pander(df, justify = c("left", "right", "right"))
```

**Limitation**: pander supports left/center/right alignment only, not decimal
alignment.

### zztab2fig Solution

```{r challenge9-zztab2fig}
df <- data.frame(
  Item = c("A", "B", "C"),
  Value1 = c(1.5, 123.45, 12.345),
  Value2 = c(0.001, 10.1, 1000.01)
)

t2f(df,
    filename = "decimal_aligned",
    align = list(
      "l",
      t2f_siunitx(table_format = "3.3"),
      t2f_siunitx(table_format = "4.2")
    ))
```

**Advantage**: zztab2fig uses siunitx for true decimal alignment.

---

## Challenge 10: Multi-Page Tables

**Task**: Create a table that spans multiple pages with repeated headers.

### pander Solution

```{r challenge10-pander}
df <- mtcars[rep(1:32, 3), ]  # 96 rows

# pander splits wide tables but not long tables
# For long tables, output continues without repeated headers
pander(df)
```

**Limitation**: pander splits wide tables horizontally but does not handle
multi-page tables with repeated headers.

### zztab2fig Solution

```{r challenge10-zztab2fig}
df <- mtcars[rep(1:32, 3), ]  # 96 rows

t2f(df,
    filename = "multipage",
    caption = "Complete mtcars Dataset (Repeated)",
    longtable = TRUE)
```

**Advantage**: zztab2fig uses LaTeX longtable for proper page breaks with
repeated headers.

---

## Challenge 11: Journal-Specific Formatting

**Task**: Format table according to NEJM (New England Journal of Medicine)
style guidelines.

### pander Solution

```{r challenge11-pander}
df <- data.frame(
  Characteristic = c("Age", "Male sex", "BMI"),
  Treatment = c("65.2 (8.4)", "58%", "27.3 (4.1)"),
  Placebo = c("64.8 (8.1)", "56%", "27.1 (3.9)")
)

# pander has no journal-specific themes
# Manual formatting required
panderOptions("table.style", "rmarkdown")
pander(df, caption = "Baseline Characteristics")
```

### zztab2fig Solution

```{r challenge11-zztab2fig}
df <- data.frame(
  Characteristic = c("Age", "Male sex", "BMI"),
  Treatment = c("65.2 (8.4)", "58%", "27.3 (4.1)"),
  Placebo = c("64.8 (8.1)", "56%", "27.1 (3.9)")
)

t2f(df,
    filename = "nejm_table",
    caption = "Baseline Characteristics",
    theme = "nejm")
```

**Advantage**: zztab2fig includes built-in themes for NEJM, APA, and Nature
journals.

---

## Challenge 12: Model Comparison Table

**Task**: Display multiple regression models side by side with significance
stars.

### pander Solution

```{r challenge12-pander}
m1 <- lm(mpg ~ cyl, data = mtcars)
m2 <- lm(mpg ~ cyl + hp, data = mtcars)
m3 <- lm(mpg ~ cyl + hp + wt, data = mtcars)

# pander displays one model at a time
pander(m1)
pander(m2)
pander(m3)

# Side-by-side comparison requires manual construction
```

**Limitation**: pander has no built-in model comparison functionality.

### zztab2fig Solution

```{r challenge12-zztab2fig}
m1 <- lm(mpg ~ cyl, data = mtcars)
m2 <- lm(mpg ~ cyl + hp, data = mtcars)
m3 <- lm(mpg ~ cyl + hp + wt, data = mtcars)

t2f_regression(
  Model1 = m1,
  Model2 = m2,
  Model3 = m3,
  stars = TRUE,
  filename = "model_comparison"
)
```

**Advantage**: zztab2fig provides `t2f_regression()` for publication-ready
model comparison tables.

---

## Challenge 13: Figure Placement in LaTeX

**Task**: Include a table as a float with specific positioning.

### pander Solution

```{r challenge13-pander}
df <- mtcars[1:5, 1:4]

# pander outputs markdown - LaTeX float control not applicable
pander(df)

# For LaTeX output, user must manually wrap in figure environment
```

**Limitation**: pander targets markdown; LaTeX float control requires manual
intervention.

### zztab2fig Solution

```{r challenge13-zztab2fig, results='asis'}
df <- mtcars[1:5, 1:4]

# Generate cropped PDF
t2f(df, filename = "float_demo", sub_dir = "tables")

# Include with float positioning
t2f_include("tables/float_demo",
            caption = "Sample Data",
            label = "fig:sample",
            position = "htbp")
```

**Advantage**: zztab2fig provides helper functions for all LaTeX placement
options (float, inline, margin, wrap).

---

## Challenge 14: Combining Multiple Features

**Task**: Create a complex table with footnotes, spanning headers, multi-row
cells, and NEJM styling.

### pander Solution

```{r challenge14-pander}
# This combination is not achievable in pander
# Would require extensive manual LaTeX coding
```

**Limitation**: pander cannot combine these advanced features.

### zztab2fig Solution

```{r challenge14-zztab2fig}
df <- data.frame(
  Endpoint = c("Primary", "Primary", "Secondary", "Secondary"),
  Timepoint = c("Week 26", "Week 52", "Week 26", "Week 52"),
  N = c(245, 232, 245, 232),
  Difference = c(-0.42, -0.58, -0.28, -0.35),
  P = c("0.008", "0.002", "0.045", "0.018")
)

# Add footnote markers
df$Difference <- sapply(seq_len(nrow(df)), function(i) {
  p <- as.numeric(df$P[i])
  if (p < 0.01) t2f_mark(as.character(df$Difference[i]), 2, "symbol")
  else if (p < 0.05) t2f_mark(as.character(df$Difference[i]), 1, "symbol")
  else as.character(df$Difference[i])
})

hdr <- t2f_header_above(" " = 2, "Results" = 3)

fn <- t2f_footnote(
  general = "Negative values favor treatment.",
  symbol = c("p < 0.05", "p < 0.01")
)

t2f(df,
    filename = "complex_table",
    caption = "Efficacy Results by Endpoint and Timepoint",
    caption_short = "Efficacy Results",
    header_above = hdr,
    collapse_rows = t2f_collapse_rows(1, valign = "top"),
    footnote = fn,
    theme = "nejm")
```

---

# Summary Comparison

| Feature | pander | zztab2fig |
|---------|:------:|:---------:|
| Basic tables | Yes | Yes |
| Column alignment | L/C/R | L/C/R + decimal |
| Captions | Yes | Yes + short captions |
| Cross-reference labels | No | Yes |
| Table footnotes | No | Yes (4 notation types) |
| Spanning headers | No | Yes |
| Multi-row cells | No | Yes |
| Multi-page tables | No | Yes (longtable) |
| Journal themes | No | Yes (NEJM, APA, Nature) |
| Model comparison | No | Yes |
| LaTeX float control | No | Yes |
| Margin placement | No | Yes |
| Output formats | Markdown | PDF, PNG, SVG, TEX |

## When to Use Each Package

**Choose pander when:**

- Working in R Markdown targeting HTML/Word output
- Need quick inline display of many R object types
- Document will be processed through Pandoc
- Simple tables without advanced formatting

**Choose zztab2fig when:**

- Creating publication-ready tables for journals
- Need footnotes, spanning headers, or multi-row cells
- Working with LaTeX documents
- Require decimal alignment
- Need consistent journal-specific styling
- Creating cropped PDF tables for inclusion in documents
