---
title: "Quickstart Guide: zzt2f (Typst Backend)"
author: "Ronald G. Thomas"
output:
  pdf_document:
    latex_engine: xelatex
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zzt2f (Typst Backend)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "45%",
  fig.align = "center"
)

has_typst <- requireNamespace("tinytable", quietly = TRUE) &&
  nzchar(Sys.which("typst"))

output_dir <- tempdir()

include_table <- function(filename) {
  img_path <- file.path(
    output_dir, paste0(filename, ".png")
  )
  if (!file.exists(img_path)) return(invisible(NULL))
  knitr::include_graphics(img_path)
}
```

This quickstart covers `zzt2f()`, the Typst-based alternative to
`t2f()`. Where `t2f()` requires a full LaTeX distribution (pdflatex
+ pdfcrop + ImageMagick), `zzt2f()` needs only two lightweight
dependencies: the tinytable R package and the Typst CLI (~40 MB
binary). Typst auto-sizes pages, so no cropping step is needed.

Both functions share the same theme system. The key difference is
the compilation backend: LaTeX vs Typst.


## Installation

```{r install, eval=FALSE}
# R package
install.packages("tinytable")

# Typst CLI
# macOS:   brew install typst
# Linux:   curl -fsSL https://typst.community | sh
# Windows: winget install --id Typst.Typst
```

```{r load}
library(zztab2fig)
```

Verify that both dependencies are available:

```{r check-deps, eval=has_typst}
check_typst_deps()
```


## Basic Usage

Convert a data frame to a table image:

```{r basic, eval=has_typst}
zzt2f(
  mtcars[1:6, 1:4],
  filename = "basic_table",
  sub_dir = output_dir,
  format = "png"
)
```

```{r basic-display, echo=FALSE, eval=has_typst}
include_table("basic_table")
```

By default, `zzt2f()` produces a PDF. Here we use `format = "png"`
for display. The output is a single file with no intermediate
artifacts (no `.tex`, no `_full.pdf`, no cropping step).


## Output Formats

`zzt2f()` supports three output formats via the `format` argument:

```{r format-pdf, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_pdf",
      sub_dir = output_dir,
      format = "pdf")
```

```{r format-png, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_png",
      sub_dir = output_dir,
      format = "png",
      dpi = 300)
```

```{r format-png-display, echo=FALSE, eval=has_typst}
include_table("fmt_png")
```

```{r format-svg, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_svg",
      sub_dir = output_dir,
      format = "svg")
```

PNG output respects the `dpi` argument (default 300). SVG output
produces resolution-independent vector graphics suitable for web
documents.


## Journal Themes

`zzt2f()` reuses the same theme system as `t2f()`. All built-in
themes translate automatically to Typst styling:

```{r theme-nejm, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      theme = "nejm",
      filename = "nejm_table",
      sub_dir = output_dir,
      format = "png")
```

```{r theme-nejm-display, echo=FALSE, eval=has_typst}
include_table("nejm_table")
```

```{r theme-apa, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      theme = "apa",
      filename = "apa_table",
      sub_dir = output_dir,
      format = "png")
```

```{r theme-apa-display, echo=FALSE, eval=has_typst}
include_table("apa_table")
```

Available themes: `"nejm"`, `"apa"`, `"nature"`, `"lancet"`,
`"minimal"`.

The global theme set via `t2f_theme_set()` is respected by
`zzt2f()` as well:

```{r global-theme, eval=FALSE}
t2f_theme_set("nejm")
zzt2f(mtcars[1:6, 1:4], filename = "auto_themed")
```


## Captions and Footnotes

```{r caption, eval=has_typst}
fn <- t2f_footnote(
  general = "Data from Motor Trend US magazine (1974).",
  symbol = c("p < 0.05", "p < 0.01")
)

zzt2f(mtcars[1:6, 1:4],
      caption = "Motor Trend Car Road Tests",
      footnote = fn,
      filename = "captioned_table",
      sub_dir = output_dir,
      format = "png")
```

```{r caption-display, echo=FALSE, eval=has_typst, out.width="55%"}
include_table("captioned_table")
```

Footnotes from `t2f_footnote()` are translated to plain-text notes
below the table. General, numbered, alphabetic, and symbol notation
styles are all supported.


## Spanning Headers

Group columns under spanning headers using `t2f_header_above()`:

```{r header, eval=has_typst}
hdr <- t2f_header_above(
  " " = 1,
  "Performance" = 2,
  "Engine" = 1
)

zzt2f(mtcars[1:6, 1:4],
      header_above = hdr,
      filename = "header_table",
      sub_dir = output_dir,
      format = "png")
```

```{r header-display, echo=FALSE, eval=has_typst}
include_table("header_table")
```


## Column Alignment

Alignment is auto-detected (numeric columns right-aligned, others
left-aligned) or set explicitly:

```{r alignment, eval=has_typst}
zzt2f(iris[1:6, ],
      align = c("r", "r", "r", "r", "l"),
      filename = "aligned_table",
      sub_dir = output_dir,
      format = "png")
```

```{r alignment-display, echo=FALSE, eval=has_typst, out.width="60%"}
include_table("aligned_table")
```

A single character applies to all columns:

```{r align-center, eval=FALSE}
zzt2f(mtcars[1:6, 1:4], align = "c")
```


## Row Stripe Color Override

Override the theme's stripe color with `scolor`:

```{r scolor, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      scolor = "blue!15",
      filename = "blue_stripes",
      sub_dir = output_dir,
      format = "png")
```

```{r scolor-display, echo=FALSE, eval=has_typst}
include_table("blue_stripes")
```

LaTeX color syntax (`"blue!15"`, `"gray!5"`) is automatically
translated to hex. Hex values (`"#E6E6FF"`) pass through unchanged.


## Working with Statistical Objects

`zzt2f()` operates on data frames only. For statistical objects,
extract the data frame first, then pass it to `zzt2f()`:

```{r stats-workflow, eval=has_typst}
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)
coef_df <- broom::tidy(model, conf.int = TRUE)
coef_df$p.value <- format_pvalue(coef_df$p.value)

zzt2f(coef_df,
      caption = "Linear Model Coefficients",
      theme = "nejm",
      filename = "lm_typst",
      sub_dir = output_dir,
      format = "png")
```

```{r stats-display, echo=FALSE, eval=has_typst, out.width="75%"}
include_table("lm_typst")
```

This two-step pattern (tidy then render) provides full control over
column selection, rounding, and formatting before the table is
generated.


## Quick Reference

### zzt2f() Parameters

| Parameter | Description | Default |
|:----------|:------------|:--------|
| `filename` | Base name (no extension) | deparsed input |
| `sub_dir` | Output directory | `"analysis/figures"` |
| `format` | `"pdf"`, `"png"`, or `"svg"` | `"pdf"` |
| `dpi` | PNG resolution | 300 |
| `caption` | Table caption | NULL |
| `align` | Column alignment (l/c/r) | NULL (auto) |
| `theme` | Theme name or object | NULL |
| `scolor` | Row stripe color override | NULL |
| `footnote` | `t2f_footnote()` object | NULL |
| `header_above` | `t2f_header_above()` object | NULL |
| `verbose` | Print progress messages | FALSE |

### Parameters from t2f() Not Available in zzt2f()

These are LaTeX-specific and have no Typst equivalent:

- `extra_packages`, `document_class`
- `caption_short`, `label`
- `longtable`
- `crop`, `crop_margin`
- `collapse_rows`

### Choosing Between t2f() and zzt2f()

| | `t2f()` | `zzt2f()` |
|:-----|:--------|:----------|
| **Backend** | LaTeX (kableExtra) | Typst (tinytable) |
| **Dependencies** | pdflatex + pdfcrop + ImageMagick | Typst CLI (~40 MB) |
| **Output formats** | PDF (+ PNG/SVG via convert) | PDF, PNG, SVG natively |
| **S3 dispatch** | Yes (lm, glm, rms, ...) | No (data frames only) |
| **Cropping** | Required (pdfcrop) | Not needed (auto-sized) |
| **Decimal alignment** | siunitx S columns | Not yet supported |
| **Multi-page tables** | longtable | Not yet supported |
| **Collapse rows** | multirow | Not yet supported |

Use `t2f()` when you need S3 dispatch, siunitx decimal alignment,
or multi-page tables. Use `zzt2f()` when you want fast compilation
with minimal system dependencies, or when you need native PNG/SVG
output without ImageMagick.


## System Requirements

- R >= 4.1 (for native pipe `|>`)
- tinytable R package
- Typst CLI (>= 0.11)

Check your setup:

```{r check-deps-end, eval=FALSE}
check_typst_deps()
```
