---
title: "Quickstart Guide: zzt2f (Typst Backend)"
author: "Ronald G. Thomas"
output:
  pdf_document:
    latex_engine: xelatex
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zzt2f (Typst Backend)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "45%",
  fig.align = "center"
)

has_typst <- requireNamespace("tinytable", quietly = TRUE) &&
  nzchar(Sys.which("typst"))

output_dir <- tempdir()

include_table <- function(filename) {
  img_path <- file.path(
    output_dir, paste0(filename, ".png")
  )
  if (!file.exists(img_path)) return(invisible(NULL))
  knitr::include_graphics(img_path)
}
```

This quickstart covers `zzt2f()`, the Typst-based alternative to
`t2f()`. Where `t2f()` requires a full LaTeX distribution (pdflatex
+ pdfcrop + ImageMagick), `zzt2f()` needs only two lightweight
dependencies: the tinytable R package and the Typst CLI (~40 MB
binary). Typst auto-sizes pages, so no cropping step is needed.

Both functions share the same theme system. The key difference is
the compilation backend: LaTeX vs Typst.


## Installation

```{r install, eval=FALSE}
# R package
install.packages("tinytable")

# Typst CLI
# macOS:   brew install typst
# Linux:   curl -fsSL https://typst.community | sh
# Windows: winget install --id Typst.Typst
```

```{r load}
library(zztab2fig)
```

Verify that both dependencies are available:

```{r check-deps, eval=has_typst}
check_typst_deps()
```


## Basic Usage

Convert a data frame to a table image:

```{r basic, eval=has_typst}
zzt2f(
  mtcars[1:6, 1:4],
  filename = "basic_table",
  sub_dir = output_dir,
  format = "png"
)
```

```{r basic-display, echo=FALSE, eval=has_typst}
include_table("basic_table")
```

By default, `zzt2f()` produces a PDF. Here we use `format = "png"`
for display. The output is a single file with no intermediate
artifacts (no `.tex`, no `_full.pdf`, no cropping step).


## Output Formats

`zzt2f()` supports three output formats via the `format` argument:

```{r format-pdf, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_pdf",
      sub_dir = output_dir,
      format = "pdf")
```

```{r format-png, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_png",
      sub_dir = output_dir,
      format = "png",
      dpi = 300)
```

```{r format-png-display, echo=FALSE, eval=has_typst}
include_table("fmt_png")
```

```{r format-svg, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      filename = "fmt_svg",
      sub_dir = output_dir,
      format = "svg")
```

PNG output respects the `dpi` argument (default 300). SVG output
produces resolution-independent vector graphics suitable for web
documents.


## Journal Themes

`zzt2f()` reuses the same theme system as `t2f()`. All built-in
themes translate automatically to Typst styling:

```{r theme-nejm, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      theme = "nejm",
      filename = "nejm_table",
      sub_dir = output_dir,
      format = "png")
```

```{r theme-nejm-display, echo=FALSE, eval=has_typst}
include_table("nejm_table")
```

```{r theme-apa, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      theme = "apa",
      filename = "apa_table",
      sub_dir = output_dir,
      format = "png")
```

```{r theme-apa-display, echo=FALSE, eval=has_typst}
include_table("apa_table")
```

Available themes: `"nejm"`, `"apa"`, `"nature"`, `"lancet"`,
`"minimal"`.

The global theme set via `t2f_theme_set()` is respected by
`zzt2f()` as well:

```{r global-theme, eval=FALSE}
t2f_theme_set("nejm")
zzt2f(mtcars[1:6, 1:4], filename = "auto_themed")
```


## Captions and Footnotes

```{r caption, eval=has_typst}
fn <- t2f_footnote(
  general = "Data from Motor Trend US magazine (1974).",
  symbol = c("p < 0.05", "p < 0.01")
)

zzt2f(mtcars[1:6, 1:4],
      caption = "Motor Trend Car Road Tests",
      footnote = fn,
      filename = "captioned_table",
      sub_dir = output_dir,
      format = "png")
```

```{r caption-display, echo=FALSE, eval=has_typst, out.width="55%"}
include_table("captioned_table")
```

Footnotes from `t2f_footnote()` are translated to plain-text notes
below the table. General, numbered, alphabetic, and symbol notation
styles are all supported.


## Spanning Headers

Group columns under spanning headers using `t2f_header_above()`:

```{r header, eval=has_typst}
hdr <- t2f_header_above(
  " " = 1,
  "Performance" = 2,
  "Engine" = 1
)

zzt2f(mtcars[1:6, 1:4],
      header_above = hdr,
      filename = "header_table",
      sub_dir = output_dir,
      format = "png")
```

```{r header-display, echo=FALSE, eval=has_typst}
include_table("header_table")
```


## Column Alignment

Alignment is auto-detected (numeric columns right-aligned, others
left-aligned) or set explicitly:

```{r alignment, eval=has_typst}
zzt2f(iris[1:6, ],
      align = c("r", "r", "r", "r", "l"),
      filename = "aligned_table",
      sub_dir = output_dir,
      format = "png")
```

```{r alignment-display, echo=FALSE, eval=has_typst, out.width="60%"}
include_table("aligned_table")
```

A single character applies to all columns:

```{r align-center, eval=FALSE}
zzt2f(mtcars[1:6, 1:4], align = "c")
```


## Row Stripe Color Override

Override the theme's stripe color with `scolor`:

```{r scolor, eval=has_typst}
zzt2f(mtcars[1:6, 1:4],
      scolor = "blue!15",
      filename = "blue_stripes",
      sub_dir = output_dir,
      format = "png")
```

```{r scolor-display, echo=FALSE, eval=has_typst}
include_table("blue_stripes")
```

LaTeX color syntax (`"blue!15"`, `"gray!5"`) is automatically
translated to hex. Hex values (`"#E6E6FF"`) pass through unchanged.


## Working with Statistical Objects

`zzt2f()` supports S3 dispatch for common statistical objects
(`lm`, `glm`, `anova`, `aov`, `htest`), automatically extracting,
rounding, and formatting coefficients:

```{r stats-workflow, eval=has_typst}
model <- lm(mpg ~ cyl + hp + wt, data = mtcars)

zzt2f(model,
      caption = "Linear Model Coefficients",
      theme = "nejm",
      filename = "lm_typst",
      sub_dir = output_dir,
      format = "png")
```

```{r stats-display, echo=FALSE, eval=has_typst, out.width="75%"}
include_table("lm_typst")
```

For full control over column selection, you can also use the
`include` parameter or pass a pre-formatted data frame directly.


## Model Comparison

Compare multiple models side-by-side with `zzt2f_regression()`:

```{r regression-comparison, eval=has_typst}
m1 <- lm(mpg ~ cyl, data = mtcars)
m2 <- lm(mpg ~ cyl + hp, data = mtcars)
m3 <- lm(mpg ~ cyl + hp + wt, data = mtcars)

zzt2f_regression(
  Model1 = m1, Model2 = m2, Model3 = m3,
  stars = TRUE,
  filename = "model_comparison",
  sub_dir = output_dir,
  format = "png"
)
```

```{r regression-display, echo=FALSE, eval=has_typst, out.width="75%"}
include_table("model_comparison")
```

A separator line is automatically inserted between coefficient
rows and summary statistics (N, R-squared, Adj. R-squared).


## Helper Functions

The `zzt2f()` pipeline supports a family of `zzt2f_*` helpers
that parallel the LaTeX-side `t2f_*` helpers. These return small
spec objects or Typst source fragments. The sections below show
representative usage.

### Cell-level formatting

Formatting specs are consumed by `zzt2f()` via its `formats`
parameter, which accepts a single spec or a list of specs.

```{r styling-demo, eval=FALSE}
library(zztab2fig)

df <- mtcars[1:4, 1:4]

zzt2f(
  df,
  filename = 'styled_demo',
  formats = list(
    zzt2f_bold_col('mpg'),
    zzt2f_italic_col('hp'),
    zzt2f_color_row(c(1, 3), background = 'gray!10'),
    zzt2f_highlight(
      function(x) as.numeric(x) > 20,
      background = 'yellow!30'
    ),
    zzt2f_decimal(c('mpg', 'disp'), digits = 1)
  )
)
```

`zzt2f_format()` is the general constructor; the other five are
thin wrappers for common cases.

### Collapsed rows and spanning headers

`zzt2f_collapse_rows()` visually merges consecutive identical
values in the specified columns; `zzt2f_header_above()` delegates
to `t2f_header_above()` and is consumed via `header_above`.

```{r structure-demo, eval=FALSE}
df <- data.frame(
  Group    = rep(c('A', 'B'), each = 3),
  Subgroup = rep(c('x', 'y', 'z'), 2),
  Value    = 1:6
)

zzt2f(
  df,
  filename = 'collapsed',
  collapse_rows = zzt2f_collapse_rows('Group'),
  header_above = zzt2f_header_above(
    ' ' = 1, 'Identifier' = 2
  )
)
```

### Footnotes, marks, and cross-references

`zzt2f_footnote()` is an alias for `t2f_footnote()`; the spec is
backend-agnostic. `zzt2f_mark()` inserts Typst superscript
markers; `zzt2f_ref()` returns a Typst cross-reference string
(`@label`).

```{r refs-demo, eval=FALSE}
df <- mtcars[1:3, 1:3]
df$mpg[1] <- zzt2f_mark(as.character(df$mpg[1]), 1, 'symbol')

zzt2f(
  df,
  filename = 'fn_demo',
  footnote = zzt2f_footnote(
    general = 'Source: mtcars (Motor Trend, 1974).',
    symbol = c('Leading observation.')
  )
)

ref_str <- zzt2f_ref('tab-fn-demo', cat = FALSE)
```

### Batch rendering

```{r batch-demo, eval=FALSE}
tables <- list(
  cars   = mtcars[1:5, 1:3],
  irises = iris[1:5,   1:3]
)
zzt2f_batch(tables, sub_dir = 'figures', theme = 'nejm')

# Per-table overrides
specs <- list(
  zzt2f_batch_spec(mtcars[1:3, 1:3], 'mt_small'),
  zzt2f_batch_spec(iris[1:3,   1:3], 'ir_small',
                   theme = 'apa')
)
zzt2f_batch_advanced(specs, sub_dir = 'figures')
```

### Fitted models

`zzt2f_tidy()` routes a `broom::tidy()` output through
`zzt2f()`; `zzt2f_coef()` is an inline wrapper for use in R
Markdown; `zzt2f_rms_compare()` builds a side-by-side comparison
of `rms` models.

```{r stat-demo, eval=FALSE}
m <- lm(mpg ~ cyl + hp, data = mtcars)
zzt2f_tidy(m, filename = 'lm_tidy', digits = 2)
zzt2f_coef(m, caption = 'Model coefficients.')
```

### Embedding a rendered table in a Typst document

The `zzt2f_include*()` helpers emit Typst source for embedding a
previously rendered PDF into a host `.typ` document. They are
invoked for their side effect (print-and-return) in R Markdown
chunks, or captured via `cat = FALSE`.

```{r include-demo, eval=FALSE}
zzt2f(mtcars[1:3, 1:3], filename = 'inner')

zzt2f_include(
  'inner',
  caption = 'Demo embedded table.',
  label = 'tab-inner',
  cat = FALSE
)
#> "#align(center)[#figure(image(\"inner.pdf\", width: 100%), ..."

# Side-by-side
zzt2f_include_sidebyside(
  'fig/a', 'fig/b',
  main_caption = 'Comparison',
  cat = FALSE
)
```

Note that Typst refuses paths outside its project root, so place
the host `.typ` file in the same directory as (or above) the
referenced PDFs.

### Preamble directives

Programmatic emitters for `.typ` preamble lines:

```{r preamble-demo, eval=FALSE}
preamble <- c(
  zzt2f_page(paper = 'a4',
             margin = list(x = '2cm', y = '2cm')),
  zzt2f_textlang('en'),
  zzt2f_font(main_font = 'Libertinus Serif')
)
```

These are intentionally thin; most users will write preamble
directives directly in their `.typ` source.

### Advanced: source-level post-processing

For users who want to transform the generated Typst source
before compilation, pass `compile = FALSE` to `zzt2f()`:

```{r source-demo, eval=FALSE}
src <- zzt2f(
  mtcars[1:3, 1:3],
  filename = 'raw_demo',
  compile = FALSE
)
# src is a character vector of .typ lines with attributes
# recording filename / sub_dir / format.

src <- c('// Added by hand', src)
zzt2f_compile(src, filename = 'raw_demo')
```


## Quick Reference

### zzt2f() Parameters

| Parameter | Description | Default |
|:----------|:------------|:--------|
| `filename` | Base name (no extension) | deparsed input |
| `sub_dir` | Output directory | `"analysis/figures"` |
| `format` | `"pdf"`, `"png"`, or `"svg"` | `"pdf"` |
| `dpi` | PNG resolution | 300 |
| `caption` | Table caption | NULL |
| `align` | Column alignment (l/c/r) | NULL (auto) |
| `theme` | Theme name or object | NULL |
| `scolor` | Row stripe color override | NULL |
| `footnote` | `t2f_footnote()` object | NULL |
| `header_above` | `t2f_header_above()` object | NULL |
| `verbose` | Print progress messages | FALSE |

### New parameters added by the helper set

| Parameter | Description |
|:----------|:------------|
| `formats` | A `zzt2f_format` spec or list of specs. |
| `collapse_rows` | A `zzt2f_collapse_rows()` spec. |
| `compile` | If `FALSE`, return Typst source instead of compiling. |

### Parameters from t2f() still without a Typst equivalent

- `extra_packages`, `document_class`: preamble is built by
  `zzt2f_page()` / `zzt2f_font()` / `zzt2f_textlang()` or by the
  user directly.
- `caption_short`: Typst `#figure` has no short-caption concept.
- `crop`, `crop_margin`: Typst auto-sizes the page; cropping is
  unnecessary.
- `longtable`: multi-page tables are not yet wired through
  `zzt2f()` although Typst itself supports them natively.

### Choosing Between t2f() and zzt2f()

| | `t2f()` | `zzt2f()` |
|:-----|:--------|:----------|
| **Backend** | LaTeX (kableExtra) | Typst (tinytable) |
| **Dependencies** | pdflatex + pdfcrop + ImageMagick | Typst CLI (~40 MB) |
| **Output formats** | PDF (+ PNG/SVG via convert) | PDF, PNG, SVG natively |
| **S3 dispatch** | Yes (lm, glm, rms, ...) | Yes (lm, glm, anova, htest) |
| **Cropping** | Required (pdfcrop) | Not needed (auto-sized) |
| **Decimal alignment** | siunitx S columns | `zzt2f_siunitx()` via Typst `metro` (partial) |
| **Multi-page tables** | longtable | Not yet wired through `zzt2f()` |
| **Collapse rows** | multirow | `zzt2f_collapse_rows()` (cell-emptying) |
| **Cell formatting** | `t2f_format()` family | `zzt2f_format()` family |
| **Figure inclusion** | `t2f_include*()` family | `zzt2f_include*()` family |
| **Batch rendering** | `t2f_batch*()` family | `zzt2f_batch*()` family |
| **Stat helpers** | `t2f_tidy`, `t2f_coef`, `t2f_rms_compare` | `zzt2f_tidy`, `zzt2f_coef`, `zzt2f_rms_compare` |

Use `t2f()` when you need multi-page tables or the full
`siunitx` option set. Use `zzt2f()` when you want fast
compilation with minimal system dependencies or native PNG/SVG
output without ImageMagick. The helper APIs are now close to
symmetric across the two backends.


## System Requirements

- R >= 4.1 (for native pipe `|>`)
- tinytable R package
- Typst CLI (>= 0.11)

Check your setup:

```{r check-deps-end, eval=FALSE}
check_typst_deps()
```
